The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-06-12 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small revisions (formating of tables, shorter abstract, additional keyword 'simulation' and minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2621 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d5452c86478035ad9f6254fcb2a1090a193413b9

\title{The Power of Unbiased Recursive Partitioning:\\
A Unifying View of CTree, MOB, and GUIDE}
\Shorttitle{The Power of Unbiased Recursive Partitioning}
%Comparing Testing Strategies of Unbiased Tree Algorithms

@@

\Abstract{
A core step of every algorithm for learning regression trees is the decision if, 
how, and where to split the underlying data -- or, in other words, the selection 
of the best splitting variable from the available covariates and the corresponding 
split point. Early tree algorithms (e.g., AID, CART) employed greedy search strategies 
that directly compare all possible split points in all available covariates. However, 
subsequent research showed that this is biased towards selection of covariates with 
more potential split points. Therefore, unbiased recursive partitioning algorithms 
have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
based on statistical inference using $p$-values that are appropriately adjusted for 
the possible split points. In a second step a split point optimizing some objective 
function is selected in the chosen split variable. However, different unbiased tree 
algorithms employ different inference frameworks for computing these $p$-values and 
their relative advantages or disadvantages are not well understood, yet.

Therefore, three different popular approaches are considered here: classical categorical 
association tests (as in GUIDE), conditional inference (as in CTree), and parameter instability 
tests (as in MOB). First, these are embedded into a common inference framework encompassing
parametric model trees, in particular linear model trees. Second, it is assessed how
different building blocks from this common framework affect the power of the 
tree algorithms to select the appropriate covariates for splitting: observation-wise
goodness-of-fit measure (residuals vs.\ model scores), dichotomization of residuals/scores
at zero, and binning of possible split variables. This shows that specifically the goodness-of-fit 
measure is crucial for the power of the procedures, with model scores without dichotomization performing
much better in many scenarios.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

% In the CTree algorithm the number of possible breaks is fixed as splits can only be made at
% observed values or optionally at an average value lying between two observations. To select the
% best split point two-sample linear statistics of the same structure as in the independence test 
% are evaluated for all possible subsets of the sample space of the selected split variable.

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline 
& Fit             & Test                     & Pruning \\ 
\hline
CTree & Non-parametric  & Conditional inference    & Pre\\
MOB   & Parametric      & Score-based fluctuation  & Pre (or post with AIC/BIC)\\
GUIDE & Parametric      & Residual-based $\chi^2$  & Post (cost-complexity pruning)\\
\hline
\end{tabular}
\caption{Default combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.
\label{tab:combinations}}
\end{table}

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l l }
\hline 
& Scores       & Dichotomization  & Categorization  & Statistic\\
\hline
CTree & Model scores & --	     & -- 	       & Sum of squares\\
MOB   & Model scores & --	     & -- 	       & Maximally selected\\
GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
\hline
\end{tabular}
\caption{Testing strategies of CTree, MOB, and GUIDE with the corresponding
setting of the building blocks in the unifying framework and the type of test statistic.
\label{tab:buildingblocks}}
\end{table}

@@

The aggregated goodness-of-fit measure (or objective or loss function) in OLS estimation
is the sum of squared residuals: $\sum_{i=1}^N \ell(y_i,x_i,\beta_0,\beta_1)$ where
$$
\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
is the squared residual which is defined as
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
These residuals can also intuitively be used as the corresponding disaggregated
observation-wise goodness-of-fit measure. Another natural candidate for this is the
score function for the $i$-th observation:
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \ell}{\partial (\beta_0, \beta_1)^{f\top}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
Thus, up to a constant scaling factor of $-2$ (that could also be omitted) the first 
component of the scores is in fact the residual. 
However, it is complemented by a second component
that captures the slope effect of $x_i$. Therefore, when computing the score for all $N$ observations
this yields an $N \times 2$ score matrix whose first column corresponds to the $N$ residuals.
As this is the derivative with respect to the intercept parameter~$\beta_0$, it
captures changes in the intercept. Moreover, the second column of the score matrix contains
derivatives with respect to the slope parameter~$\beta_1$ and thus captures changes in this.

@@

\begin{table}[t!]
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$\\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{\label{tab:variables}Variables included in the data generating process 
as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_{k-1}(Z_1,Z_2)$, $k=1,2=K$.}
\end{table}

@@

\subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
\label{sec:scores}

@@

\begin{figure}
\setkeys{Gin}{width=\linewidth}
<<scores_stump, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                         "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree","mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l }
\hline
Building block               & Levels              & Transformation \\
\hline
Residuals vs.\ scores        & \textit{residuals}  & $h(Y) = r(Y,X,\hat{\beta})$\\
\quad ($h$ transformation)   & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
\hline
Dichotomization              & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
\quad (of $h$)               & \textit{no}         & $h$ without dichotomization\\
\hline
Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)^{\top}$\\
\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
& \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)^{\top}$\\
&                     & \quad indicating the potential split point\\
& \textit{lin}        & $g(Z)=Z$\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

\begin{table}[p!]
\centering
\begin{tabular}{ l l l l l}
\hline\noalign{\smallskip} 
Residuals/Scores   & Dichotomization & Categorization & \multicolumn{2}{l}{Via testing strategy}\\
\noalign{\smallskip}\hline\noalign{\smallskip}
residuals	   & yes	     & cat            & GUIDE & {(default)}\\
residuals	   & yes	     & max            & MOB   & {(with modifications)}\\
residuals	   & yes	     & lin            & CTree & {(with modifications)}\\
residuals	   & no 	     & cat            & MOB   & {(with modifications)}\\
residuals	   & no 	     & max            & MOB   & {(with modifications)}\\
residuals	   & no 	     & lin            & CTree & {(with modifications)}\\
scores  	   & yes	     & cat            & GUIDE & {(with modifications)}\\
scores  	   & yes	     & max            & MOB   & {(with modifications)}\\
scores  	   & yes	     & lin            & CTree & {(with modifications)}\\
scores  	   & no 	     & cat            & MOB   & {(with modifications)}\\
scores  	   & no 	     & max            & MOB   & {(default)}\\
scores  	   & no 	     & lin            & CTree & {(default)}\\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Testing strategies applied for all 12 factorial combinations of the building blocks.
For each combination one of the three testing strategies CTree/MOB/GUIDE is applied,
possibly after modifying the assessed variables as indicated (i.e., dichotomized or categorized).
\label{tab:combinations_appendix}}
\end{table}

>>>>

\title{The Power of Unbiased Recursive Partitioning:\\
A Unifying View of CTree, MOB, and GUIDE}
\Shorttitle{The Power of Unbiased Recursive Partitioning}

@@

\Abstract{
A core step of every algorithm for learning regression trees is the selection 
of the best splitting variable from the available covariates and the corresponding 
split point. Early tree algorithms (e.g., AID, CART) employed greedy search strategies, 
directly comparing all possible split points in all available covariates. However, 
subsequent research showed that this is biased towards selecting covariates with 
more potential split points. Therefore, unbiased recursive partitioning algorithms 
have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
based on statistical inference using $p$-values that are adjusted for the possible split 
points. In a second step a split point optimizing some objective function is selected in 
the chosen split variable. However, different unbiased tree algorithms obtain these 
$p$-values from different inference frameworks and their relative advantages or 
disadvantages are not well understood, yet.

Therefore, three different popular approaches are considered here: classical categorical 
association tests (as in GUIDE), conditional inference (as in CTree), and parameter 
instability tests (as in MOB). First, these are embedded into a common inference framework
encompassing parametric model trees, in particular linear model trees. Second, it is 
assessed how different building blocks from this common framework affect the power of the 
algorithms to select the appropriate covariates for splitting: observation-wise
goodness-of-fit measure (residuals vs.\ model scores), dichotomization of residuals/scores
at zero, and binning of possible split variables. This shows that specifically the 
goodness-of-fit measure is crucial for the power of the procedures, with model scores 
without dichotomization performing much better in many scenarios.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning,
simulation}

@@

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline\noalign{\smallskip} 
& Fit             & Test                     & Pruning \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
CTree & Non-parametric  & Conditional inference    & Pre\\
MOB   & Parametric      & Score-based fluctuation  & Pre (or post with AIC/BIC)\\
GUIDE & Parametric      & Residual-based $\chi^2$  & Post (cost-complexity pruning)\\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Default combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.
\label{tab:combinations}}
\end{table}

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l l }
\hline\noalign{\smallskip} 
& Scores       & Dich.  & Cat.  & Statistic\\
\noalign{\smallskip}\hline\noalign{\smallskip}
CTree & Model scores & --	     & -- 	       & Sum of squares\\
MOB   & Model scores & --	     & -- 	       & Maximally selected\\
GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Testing strategies of CTree, MOB, and GUIDE with the corresponding
setting of the building blocks in the unifying framework and the type of test statistic.
\label{tab:buildingblocks}}
\end{table}

@@

The aggregated goodness-of-fit measure (or objective or loss function) in OLS estimation
is the sum of squared residuals: $\sum_{i=1}^N \ell(y_i,x_i,\beta_0,\beta_1)$ where
$$
\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
is the squared residual which is defined as
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
These residuals can also intuitively be used as the corresponding disaggregated
observation-wise goodness-of-fit measure. Another natural candidate for this is the
score function for the $i$-th observation:
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \ell}{\partial (\beta_0, \beta_1)^{\top}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
Thus, up to a constant scaling factor of $-2$ (that could also be omitted) the first 
component of the scores is in fact the residual. 
However, it is complemented by a second component
that captures the slope effect of $x_i$. Therefore, when computing the score for all $N$ observations
this yields an $N \times 2$ score matrix whose first column corresponds to the $N$ residuals.
As this is the derivative with respect to the intercept parameter~$\beta_0$, it
captures changes in the intercept. Moreover, the second column of the score matrix contains
derivatives with respect to the slope parameter~$\beta_1$ and thus captures changes in this.

@@

\begin{table}[t!]
\centering
\begin{tabular}{l l l}
\hline\noalign{\smallskip} 
Name                    & Notation                   & Specification \\
\noalign{\smallskip}\hline\noalign{\smallskip} 
\textit{Variables:}     &                        & \\ 
Response                & $Y$                    & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                    & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$             & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                  & $\mathcal{U}([-1,1])$\\
Noise split variables   & $Z_2$--$Z_{10}$   & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
                        &                        & (alternating) \\
\noalign{\smallskip}\hline\noalign{\smallskip} 
\multicolumn{3}{l}{\textit{Parameters/functions:}} \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\noalign{\smallskip}\hline 
\end{tabular}
\caption{\label{tab:variables}Variables included in the data generating process 
as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_{k-1}(Z_1,Z_2)$, $k=1,2=K$.}
\end{table}

@@

\newpage
\subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
\label{sec:scores}

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\linewidth}
<<scores_stump, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                         "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree","mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

\begin{table}
\centering
\begin{tabular}{ l l l }
\hline\noalign{\smallskip} 
Building block               & Levels              & Transformation \\
\noalign{\smallskip}\hline\noalign{\smallskip} 
Residuals vs.\ scores        & \textit{residuals}  & $h(Y) = r(Y,X,\hat{\beta})$\\
($h$ transformation)         & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
\noalign{\smallskip}\hline\noalign{\smallskip} 
Dichotomization              & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
(of $h$)                     & \textit{no}         & $h$ without\\
                             &                     & \quad dichotomization\\
\noalign{\smallskip}\hline\noalign{\smallskip} 
Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)^{\top}$\\
($g$ transformation)         &                     & \quad indicating the\\
                             &                     & \quad assigned bin\\
& \textit{max}               & $g(Z)=(0,\ldots,0,1,\ldots,1)^{\top}$\\
&                            & \quad indicating the\\
&                            & \quad potential split point\\
& \textit{lin}        & $g(Z)=Z$\\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

\begin{table}[p!]
\centering
\begin{tabular}{ l l l l l}
\hline\noalign{\smallskip} 
Residuals/Scores   & Dich. & Cat. & \multicolumn{2}{l}{Via testing strategy}\\
\noalign{\smallskip}\hline\noalign{\smallskip}
residuals	   & yes	     & cat            & GUIDE & {(default)}\\
residuals	   & yes	     & max            & MOB   & {(modified)}\\
residuals	   & yes	     & lin            & CTree & {(modified)}\\
residuals	   & no 	     & cat            & MOB   & {(modified)}\\
residuals	   & no 	     & max            & MOB   & {(modified)}\\
residuals	   & no 	     & lin            & CTree & {(modified)}\\
scores  	   & yes	     & cat            & GUIDE & {(modified)}\\
scores  	   & yes	     & max            & MOB   & {(modified)}\\
scores  	   & yes	     & lin            & CTree & {(modified)}\\
scores  	   & no 	     & cat            & MOB   & {(modified)}\\
scores  	   & no 	     & max            & MOB   & {(default)}\\
scores  	   & no 	     & lin            & CTree & {(default)}\\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Testing strategies applied for all 12 factorial combinations of the building blocks.
For each combination one of the three testing strategies CTree/MOB/GUIDE is applied,
possibly after modifying the assessed variables as indicated (i.e., dichotomized or categorized).
\label{tab:combinations_appendix}}
\end{table}
